---
title: "Spring Seasonal Change at Campus Point — UCSB Lagoon Field Study"
date: "2026-05-28"
categories: [field study, ecology, data collection, ENVS 153]
description: "An 8-week field observation study tracking seasonal changes in plants, pollinators, and bird species richness at UCSB's Campus Point lagoon."
---

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(janitor)
library(here)

# read in group 6 observation data
observations_raw <- read_csv(
  here("data", "group6_observations.csv"),
  col_names = c(
    "group_number", "location_n", "location_w", "date", "week_number",
    "max_temp", "min_temp", "precip", "plant_species", "common_name_plant",
    "flower_head_count", "bird_species", "insect_species", "insect_common_name",
    "flowering_stage_wild_radish", "great_brome_seeding_percent",
    "humidity", "tadpole_presence", "abundancy_of_pollinators", "algae_coverage_percent"
  ),
  skip = 1
)

# clean and fill down week/date info
observations_clean <- observations_raw |>
  fill(group_number, location_n, location_w, date, week_number, .direction = "down") |>
  mutate(
    week_number = as.numeric(week_number),
    algae_coverage_percent = as.numeric(algae_coverage_percent),
    great_brome_seeding_percent = suppressWarnings(as.numeric(great_brome_seeding_percent)),
    flower_head_count = suppressWarnings(as.numeric(flower_head_count))
  )
```

## Overview

As part of ENVS 153 at UC Santa Barbara, our team conducted an 8-week field study tracking spring seasonal changes in vegetation, pollinators, and bird diversity at **Campus Point** — a coastal sage scrub and grassland site located next to Campus Point Beach, adjacent to UCSB's Cheadle Center restoration area.

**Team:** Izzy Ribondo, Yaasiel Leos, and Zhelin Qi

**Location:** UCSB Campus Lagoon, Goleta, CA (34°24'22.7"N, 119°50'39.8"W)

---

## Why Campus Point?

We chose this site because of its rich ecological diversity and its position at the edge of the lagoon ecosystem. It offered:

- High abundance of **Wild Radish** (*Raphanus sativus*) and **Great Brome** (*Bromus diandrus*) for tracking non-native plant changes
- Presence of birds, pollinators, and lagoon wildlife
- Proximity to the Cheadle Center coastal restoration area
- A mix of native and non-native species expected to show rapid seasonal change through spring

A native **Torrey Pine** (*Pinus torreyana*) — one of North America's rarest pine species — was also present at the site, supporting lichen growth and hummingbird nesting.

---

## Research Questions

We investigated three main questions:

**Q1:** How does native and non-native vegetation change throughout the season near the lagoon at UCSB?

**Q2:** How does the increase in flowering plants affect pollinator abundance and activity near Campus Point?

**Q3:** How did bird species richness change throughout spring at Campus Point?

---

## Methods

Every Thursday at 1:00 PM over 8 weeks, we visited the same site and recorded:

- Temperature, weather, and site conditions
- Bird species (identified with Merlin Bird ID)
- Plant species and flower/seed counts (identified with Seek and iNaturalist)
- 5-minute pollinator surveys
- Photos and timelapse documentation of site change

---

## Results

### Algae Coverage Over Time

```{r algae-plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height=4, fig.width=5}
algae_data <- observations_clean |>
  select(week_number, algae_coverage_percent) |>
  drop_na(algae_coverage_percent) |>
  distinct(week_number, .keep_all = TRUE)

ggplot(algae_data, aes(x = week_number, y = algae_coverage_percent)) +
  geom_point(size = 3, alpha = 0.8, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "white") +
  labs(x = "Observation week", y = "Algae coverage (%)") +
  theme_classic() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )
```

Algae coverage ranged from 10% in Week 1 to a peak of 27% in Week 6, showing a general increasing trend through mid-season.

---

### Pollinator Abundance Over Time

```{r pollinator-plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height=4, fig.width=5}
pollinator_data <- tibble(
  week_number = 1:8,
  pollinators_per_5_min = c(1, 4, 5, 7, 11, 9, 13, 2)
)

ggplot(pollinator_data, aes(x = week_number, y = pollinators_per_5_min)) +
  geom_point(size = 3, alpha = 0.8, color = "mediumpurple") +
  geom_smooth(method = "lm", se = FALSE, color = "white") +
  labs(x = "Observation week", y = "Pollinators observed per 5 minutes") +
  theme_classic() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )
```

Pollinator abundance peaked at **13 per 5 minutes in Week 7**, then dropped sharply to 2 in Week 8 as flowers senesced. Pollinators observed included Western Honeybees, cabbage white butterflies, and hummingbirds.

---

### Great Brome Seeding Percentage

```{r brome-plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height=4, fig.width=5}
great_brome_data <- observations_clean |>
  filter(common_name_plant == "Ripgut Brome/ Great Brome") |>
  select(week_number, great_brome_seeding_percent) |>
  drop_na(great_brome_seeding_percent)

ggplot(great_brome_data, aes(x = week_number, y = great_brome_seeding_percent)) +
  geom_point(size = 3, alpha = 0.8, color = "darkorange") +
  geom_smooth(method = "lm", se = FALSE, color = "white") +
  labs(x = "Observation week", y = "Great brome seeding (%)") +
  theme_classic() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )
```

Great Brome seeding percentage declined from 70% in Week 2 to just 1% by Week 8, marking the completion of its reproductive cycle.

---

### Flower and Seed Head Counts by Plant Species

```{r plant-plot, message=FALSE, warning=FALSE, echo=FALSE, fig.height=4, fig.width=6}
plant_count_data <- observations_clean |>
  filter(!is.na(common_name_plant)) |>
  drop_na(flower_head_count)

ggplot(plant_count_data, aes(x = week_number, y = flower_head_count, color = common_name_plant)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Observation week",
    y = "Flower or seed head count",
    color = "Plant species"
  ) +
  theme_classic() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    legend.background = element_rect(fill = "transparent", color = NA)
  )
```

Wild Radish dominated early in the season, peaking at over **436 flower/seed heads in Week 4** before declining sharply. Invasive species consistently outnumbered native species throughout the observation period.

---

## Key Takeaways

This project gave me a firsthand look at how seasonal change drives cascading ecological effects — from plant phenology to insect activity to bird diversity. Watching the site transform week by week, from lush green vegetation in early spring to dry, seed-heavy grasses by late May, made abstract ecological concepts tangible. It also reinforced the impact of invasive species: Wild Radish and Great Brome consistently dominated the site, shaping the resources available to every other organism present.
